Matrix Transposition and Symmetry: a case study on Superscalar Architectures\\

### 2024-12-10 latex/paper.tex

\title{Matrix Transposition and Symmetry in Superscalar Architectures: A Case Study\\
%{\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and
%should not be used}
%\thanks{Identify applicable funding agency here. If none, delete this.}
}

@@

In section 2, we discuss the state of the art of the two algorithms.
In section 3, we elaborate on the algorithms analyzed.
In section 4, we explain the benchmarking procedures and system information.
In section 5, we present the results. The conclusion is in section 6. Section 7 contains
the project's github link.

@@

Computing the transpose of a matrix is a well discussed computational problem. Early
work focuses on in-situ transporision, that is transposing a matrix without
relying on additional space. Remarcable contributions include Boothroyd's ACM Algorithm 302 (1966) using cyclic mapping \cite{b13}, Laflin and Brefner ACM Algorithm 380
adding fixed points arithmetic \cite{b14} and
a later revision of the two by Esko G. Cate and David W. Twigg in ACM Algorithm 513 \cite{b15} .
SIMD implementations for specific architectures are also widely adopted
in performance critical applications like cryptographic libraries, utilizing instructions like Intel's vpunpckldq and vpunpckhdq \cite{b16} and reporting improvements up to 70\% \cite{b17}.
However, vector registers are unoptimal for matrices since registers read data only sequencially, motivating research on new vector register file (VFS) designs such as diagonal registers \cite{b23}.
Another area of research includes out-of-core matrix transposition, that
is computing the transpose of a matrix that exceeds the size of the available in-core memory.
This problem focuses on optimizing the number of I/O operations and file layout since memory-memory data transfer is notably slower than math operations. Work in out-of-core transposition include W. 0. Alltop (1975) \cite{b18}, Suh \& Prasanna (2002) \cite{b19} and Krishnamoorth et al. (2004) \cite{b20}.
Recent work explores parallelization and the use of GPUs to offload algebra operations. Examples are I-Jui Sung et al \cite{b21} and Mark Harris \cite{b22}.

\iffalse
\textit{A square matrix $A$ is called symmetric if it equals its transpose} \cite{b12}.
\fi 
In constrast, while symmetric matrces are actively used in various areas of research as their
properties allow for several optimizations in various problems, an in depth case study of symmetry
checking on modern architectures is yet to be published.

@@

In this report, different implementations of \textit{matTranspose}
and \textit{checkSymm} are benchmarked, comparing implicit and explicit parallelism
techniques. In particular, the benchmarks aim to indicate cache and SIMD performance,
speedup with multiple threads and the impact of compile flags. Data will be compared visually via graphs. All benchmarks
will be tested against every matrix size of power of $2$ from $2^2$ to $2^{12}$.  \\
The algorithms for \textit{matTranspose} will now be explained.
Each algorithm takes a $N \times N$ input square matrix \textit{input}, a matrix of the same shape \textit{ouput}, and the size of the two matrices \textit{size}.
The analyzed algorithms are the following:
\begin{itemize}
\item \textit{matTranspose}: naive implementation of matTranspose, iterating over values of the input matrix by rows and columns and storing them in the output matrix with indexes swapped. See Algorithm~\ref{matTranspose} for the pseudocode.

@@

\item \textit{matTransposeCyclic}: implementation of \cite{b28}, storing the matrix as a contiguous one-dimensional array of size $N*N$. See Algorithm~\ref{matTransposeCyclic} for the pseudocode.
  \begin{algorithm}
  	\caption{matTransposeCyclic}\label{matTransposeCyclic}
  \begin{algorithmic}[1]

  	  \For{\textit{n} \textbf{ in } \textit{0..size*size-1}}
        \State $int\ \textit{i} \gets \frac{n}{size}$
        \State $int\ \textit{j} \gets n \mod size$
        \State $T[n] \gets M[size * j + i]$
        \EndFor
          \end{algorithmic}
  \end{algorithm}

@@

The algorithms for \textit{checkSymm} will now be explained. Each algorithm takes an $N \times N$ square matrix \textit{input}, a size \textit{size} and return \textit{true} if the matrix is symmetrix, \textit{false} otherwise.
The analyzed algorithms are the following:
\begin{itemize}
\item \textit{checkSymm}: naive implementation of checkSymm. It Iterates over the upper half of the matrix and testing if every value is equal to the symmetric value.
% See Algorithm~\ref{checkSymm} for the pseudocode.

@@

@@

@@

@@

@@

Github link: \href{https://github.com/San7o/parallel-computing-cpp/tree/first-assignment}{https://github.com/San7o/parallel-computing-cpp/tree/first-assignment} \\
Please read the instructions in the \textit{README.md} for building.

\begin{thebibliography}{00}
\bibitem{b1} Intel® 64 and IA-32 Architectures Software Developer’s Manual Combined Volumes: 1, 2A, 2B, 2C, 2D, 3A, 3B, 3C, 3D, and 4. Volume 1, Section 2.2.7 "SIMD Instructions"
\bibitem{b2} SIMD man page "https://man.freebsd.org/cgi/man.cgi?simd(7)"
\bibitem{b3} Git blame on freeBSD's source code, in "lib/libc/amd64/string", shows recent activity
\bibitem{b4} https://freebsdfoundation.org/blog/a-sneak-peek-simd-enhanced-string-functions-for-amd64/
\bibitem{b5} Amodei, D. Hernandez, D. AI and Compute, https://openai.com/blog/ai-and-compute
\bibitem{b6} Andrew J. Lohn, Micah Musse. AI and Compute How Much Longer Can Computing Power Drive Artificial Intelligence Progress? CSET Issue Brief
\bibitem{b7} Digitalized Apollo 11 guidance computer by Virtual AGC and MIT Museum. Code in "Apollo-11/Comanche055/PLANETERY\_INTERNAL\_ORIENTATION.agc" line 34; "Apollo-11/Luminary099/ATTITUDE\_MANEUVER\_ROUTINE.agc" line 441
\bibitem{b8} Matteo Frigo and Steven G. Johnson. "The design and implementation of FFTW3". The word "transpose" is mentioned 21 times
\bibitem{b9} M. Henriksson and O. Gustafsson, "Streaming Matrix Transposition on FPGAs Using Distributed Memories," 2023 IEEE Nordic Circuits and Systems Conference (NorCAS), Aalborg, Denmark, 2023, pp. 1-6, doi: 10.1109/NorCAS58970.2023.10305472.
\bibitem{b10} "https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2023/p1673r12.html" Accessed in november 2024
\bibitem{b11} Sheldon Axler. Linear Algebra Done Right ourth edition
16 November 2024. Section 3C Matrices, definition 3.54
\bibitem{b12} Sheldon Axler. Linear Algebra Done Right ourth edition
16 November 2024. Section 9A Bilinear Forms and Quadratic Forms, definition 9.11
\bibitem{b13} BOOTHROYD,Z. Algorithm 302: Transpose vector stored array. Go~r~m.A~M I0, ~ (1967),
292-293
\bibitem{b14} LAFLIN, S., AND BREFNER, M.A. Algorithm 380: In-situ transposition of a rectangular
matrix, Comm. ACM IS, 5 (1970), 324-326
\bibitem{b15} ESKO G. CATE and DAVID W. TWIGG ALGORITHM 513: Analysis of In-Situ Transposition (1977)
\bibitem{b16} most of the crypto functions are written in assembly usin SIMD instructions in the
linux kernel "linux/arch/x86/crypto" for exanple transpose\_4x4 in "serpent-avx2-asm\_64.S", "sm4-aesni-avx-asm\_64.S",
"cast6-avx-x86\_64-asm\_64.S" and others
\bibitem{b17} https://github.com/torvalds/linux/commit/d34a460092d857f1616e39eed7eac6f40cea2225
\bibitem{b18} W. O. Alltop, "A Computer Algorithm for Transposing Nonsquare Matrices," in IEEE Transactions on Computers, vol. C-24, no. 10, pp. 1038-1040, Oct. 1975, doi: 10.1109/T-C.1975.224124.
\bibitem{b19} Jinwoo Suh and V. K. Prasanna, "An efficient algorithm for out-of-core matrix transposition," in IEEE Transactions on Computers, vol. 51, no. 4, pp. 420-438, April 2002, doi: 10.1109/12.995452.
\bibitem{b20} Kandemir, M. \& Choudhary, A. \& Ramanujam, J. \& Arunachalam, Meenakshi. (2000). A unified framework for optimizing locality, parallelism, and communication in out-of-core computations. Parallel and Distributed Systems, IEEE Transactions on. 11. 648 - 668. 10.1109/71.877759. 
\bibitem{b21} I-Jui Sung, Juan G´omez-Luna, Josè Marìa Gonzàlez-Linares
Nicolàs Gui, Wen-Mei W. Hwu. In-Place Transposition of Rectangular Matrices on Accelerators
\bibitem{b22} Post: An Efficient Matrix Transpose in CUDA C/C++. Feb 18, 2013 By Mark Harris, https://developer.nvidia.com/blog/efficient-matrix-transpose-cuda-cc/
\bibitem{b23} Hanounik, Bedros. (2000). Diagonal Registers: Novel Vector Register File Design for High Performance and Multimedia Computing. 
\bibitem{b24} https://github.com/San7o/valFuzz
\bibitem{b25} Rotenberg, A. (1960). "A New Pseudo-Random Number Generator". Journal of the ACM. 7 (1): 75–77
\bibitem{b26} Numerical Recipes ranqd1, Chapter 7.1, An Even Quicker Generator, Eq. 7.1.6
parameters from Knuth and H. W. Lewis
\bibitem{b27} https://github.com/San7o/tenno-tl
\bibitem{b28} MARTIN F. BERMAN.  A method for Transposing a Matrix. System Developement Corporation, Santa Monica, California.
\end{thebibliography}
\vspace{12pt}

>>>>

\title{Matrix Transposition and Symmetry: a case study on Superscalar Architectures\\
%{\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and
%should not be used}
%\thanks{Identify applicable funding agency here. If none, delete this.}
}

@@

In section 2, we discuss the state of the art of the two algorithms.
In section 3, we elaborate on the algorithms analyzed.
In section 4, we explain the benchmarking procedures and system information.
In section 5, we present the results. The conclusion is in section 6. Section 7 contains
the project's github link

@@

Computing the transpose of a matrix is a well discussed computational problem. Early
work focuses on in-situ transporision, that is transposing a matrix without
relying on additional space. Implementations were often based on the cyclic
structure of the transposition mapping: values of a matrix of size $N$ can be stored
in a single contiguous array or size $N$*$N$ and indexed with $row*\#rows+ column$.
Remarcable contributions include Boothroyd's ACM Algorithm 302 (1966) using cyclic mapping \cite{b13}, Laflin and Brefner ACM Algorithm 380
adding fixed points arithmetic \cite{b14} and
a later revision of the two by Esko G. Cate and David W. Twigg in ACM Algorithm 513 \cite{b15} .
SIMD implementations for specific architectures are also widely adopted
in performance critical applications like cryptographic libraries, utilizing instructions like Intel's vpunpckldq and vpunpckhdq \cite{b16} and reporting improvements up to 70\% \cite{b17}.
However, vector registers are unoptimal for matrices since registers read data only sequencially, motivating research on new vector register file (VFS) designs such as diagonal registers \cite{b23}.
Another area of research includes out-of-core matrix transposition, that
is computing the transpose of a matrix that exceeds the size of the available in-core memory.
This problem focuses on optimizing the number of I/O operations and file layout since memory-memory data transfer is notably slower than math operations. Work in out-of-core transposition include W. 0. Alltop (1975) \cite{b18}, Suh \& Prasanna (2002) \cite{b19} and Krishnamoorth et al. (2004) \cite{b20}.
Recent work explores parallelization and the use of GPUs to offload algebra operations. Examples are I-Jui Sung et al \cite{b21} and Mark Harris \cite{b22}.

\iffalse
\textit{A square matrix $A$ is called symmetric if it equals its transpose} \cite{b12}.
\fi 
While symmetric matrces are actively used in various areas of research as their
properties allow for several optimizations in various problems, an in depth case study of symmetry
checking on modern architectures is yet to be published.

@@

In this report different implementations of \textit{matTranspose}
and \textit{checkSymm} are benchmarked, comparing implicit and explicit parallelism
techniques. In particular, the benchmarks aim to indicate cache and SIMD performance differences,
speedup with multiple threads and the impact of compile flags optimizations. Data will be compared visually via graphs. All benchmarks
will be tested against every matrix size from $2^2$ to $2^{12}$, we will assume the size value is a multiple of 4 for ease of implementation.  \\
The algorithms for \textit{matTranspose} will now be explained.
Each algorithm takes a $N \times N$ input square matrix \textit{input}, a matrix of the same shape \textit{ouput}, and the size of the two matrices \textit{size}.
The analyzed algorithms are the following:
\begin{itemize}
\item \textit{matTranspose}: naive implementation of matTranspose, iterating over values of the input matrix and storing them in the output matrix with indexes of rows and columns swapped. See Algorithm~\ref{matTranspose} for the pseudocode.

@@

\item \textit{matTransposeCyclic}: implementation of \cite{b13}, storing the matrix as a contiguous one-dimensional array of size $N*N$. See Algorithm~\ref{matTransposeCyclic} for the pseudocode.
  \begin{algorithm}
  	\caption{matTransposeCyclic}\label{matTransposeCyclic}
  \begin{algorithmic}[1]

  	  \For{\textit{n} \textbf{ in } \textit{0..size*size-1}}
        \State $int\ \textit{i} \gets \frac{n}{size}$
        \State $int\ \textit{j} \gets n \mod N$
        \State $T[n] \gets M[size * j + i]$
        \EndFor
          \end{algorithmic}
  \end{algorithm}

@@

The algorithms for \textit{checkSymm} will now be explained. Each algorithm take an $N \times N$ square matrix \textit{input}, a size \textit{size} and return true if the matrix is symmetrix, false otherwise.
The analyzed algorithms are the following:
\begin{itemize}
\item \textit{checkSymm}: naive implementation of checkSymm. It Iterates over the upper half of the matrix and testing if every value is equal to the symmetric value.
% See Algorithm~\ref{checkSymm} for the pseudocode.

@@

\iffalse
Remember: write
- what you did (contribution): high level overview like "hey, i want to test different vectorization ond parallelization implementations"
- how you did it (methodology): we benchmarked algorithm A ... (all the pseudocodes). Whay we mesured

@@

\iffalse
The experiments were conducted on a cluster equipped with 64 Intel(R) Xeon(R) Gold 6140M CPU with 2.30GHz, 32K of L1d cache, 32K L1i cache, 1025K L2 cache, 25344K L3 cache with cache lines of 64 bits, 1 thread per core, 4GB of DDR4 RAM running Linux 3.10.
The algorithms were benchmarked using valFuzz \cite{b24}. The set of benchmarks were compiled using gcc-9.1.0 with flags "-fopenmp", "-fopenmp-simd", 4 set of benchmarks are compiled with either no optimizations, -O1, -O2 and "-O3 -Ofast -march=native".
To obtain results with statistical significance, each benchmark has been executed 100 times collecting the following data concerning wall-clock execution time: minimum, maximum, median, arithmetic mean, standard deviation, Q1, Q3.
To prevent interference between the benchmarks, the cache is cleared by initializing an array twice as big as the L3 cache size.
To ensure fair comparison, the same input matrix is provided to all algorithms.
The input matrix is initialized with an uniform random distribution of floats ranging from -1 to 1 generated using a Linear Congruential Generator (LGC) \cite{b25} with the parameters $a=1664525$, $c = 1013904223$, $m=2^{32}$ \cite{b26} and seed $1337$.
To save time, generation of random numbers is done at compile time using tenno-tl \cite{b27}. Additional algorithms and data structures were used from tenno-tl including ranges and arrays.
\fi

@@

Mamma mia ma quanto e' veloce sto asesmbly,
buu al compilatore che non ci azzecca per bene. 
Anche ciclico è nammerda, invece l'unrolling
è figo.
In generale abbiamo capito
cose ed il cluster è pazzesco e le differenze
sono notevoli, evviva i soldi.

Ok fatto un po' meglio:

@@

\iffalse
The results are now presented, matTranspose will be analyzed first.
The roofline model in figure~\ref{fig:transpose_benchmarks} (d) shows that the base algorithm is memory bounded so improvement over accessing memory efficiently will be rewarded. Starting from the algorithms without any parallelism depicted in figure~\ref{fig:transpose_base}, we notice \textit{matTransposeCyclic} to be constantly
slower that the others, while \textit{matTransposeColumns} shows little to no difference compared to the base algorithm with the same compiler flags. The performance difference between the other algorithms increases as more aggressive compiler optimizations are enabled. \textit{checkSymmHalf} becomes slower than the base line with the same optimizations, when they are enabled. 
Taking into consideration implicit parallelism techniques in figure~\ref{fig:transpose_benchmarks} (a), the faster algorithm is \textit{matTransposeIntrinsicCyclic} from size smaller than $10^2$, getting surpassed by \textit{matTransposeIntrinsic} and \textit{matTransposeUnrollingOuter} in larger matrices. \textit{matTransposeCyclicUnrolled} is the slower for all sizes. Looking at explicit parallelism in figure-\ref{fig:transpose_benchmarks} (b), benchmarks with giher number of threads take more time than benchmarks with fewer threads until around size $10^2$ to $10^3$ where the trend is inverted. When analysing schedulers, static is slower for most matrix sixzes but all algorithms converge to about the same timing in the bigger input. 
\fi

@@

Github link: \href{https://github.com/San7o/parallel-computing-cpp}{https://github.com/San7o/parallel-computing-cpp} \\
Please read the instructions in the \textit{README.md} for building.

\begin{thebibliography}{00}
\bibitem{b1} Intel® 64 and IA-32 Architectures Software Developer’s Manual Combined Volumes: 1, 2A, 2B, 2C, 2D, 3A, 3B, 3C, 3D, and 4. Volume 1, Section 2.2.7 "SIMD Instructions"
\bibitem{b2} SIMD man page "https://man.freebsd.org/cgi/man.cgi?simd(7)"
\bibitem{b3} Git blame on freeBSD's source code, in "lib/libc/amd64/string", shows recent activity
\bibitem{b4} https://freebsdfoundation.org/blog/a-sneak-peek-simd-enhanced-string-functions-for-amd64/
\bibitem{b5} Amodei, D. Hernandez, D. AI and Compute, https://openai.com/blog/ai-and-compute
\bibitem{b6} Andrew J. Lohn, Micah Musse. AI and Compute How Much Longer Can Computing Power Drive Artificial Intelligence Progress? CSET Issue Brief
\bibitem{b7} Digitalized Apollo 11 guidance computer by Virtual AGC and MIT Museum. Code in "Apollo-11/Comanche055/PLANETERY\_INTERNAL\_ORIENTATION.agc" line 34; "Apollo-11/Luminary099/ATTITUDE\_MANEUVER\_ROUTINE.agc" line 441
\bibitem{b8} Matteo Frigo and Steven G. Johnson. "The design and implementation of FFTW3". The word "transpose" is mentioned 21 times
\bibitem{b9} M. Henriksson and O. Gustafsson, "Streaming Matrix Transposition on FPGAs Using Distributed Memories," 2023 IEEE Nordic Circuits and Systems Conference (NorCAS), Aalborg, Denmark, 2023, pp. 1-6, doi: 10.1109/NorCAS58970.2023.10305472.
\bibitem{b10} "https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2023/p1673r12.html" Accessed in november 2024
\bibitem{b11} Sheldon Axler. Linear Algebra Done Right ourth edition
16 November 2024. Section 3C Matrices, definition 3.54
\bibitem{b12} Sheldon Axler. Linear Algebra Done Right ourth edition
16 November 2024. Section 9A Bilinear Forms and Quadratic Forms, definition 9.11
\bibitem{b13} BOOTHROYD,Z. Algorithm 302: Transpose vector stored array. Go~r~m.A~M I0, ~ (1967),
292-293
\bibitem{b14} LAFLIN, S., AND BREFNER, M.A. Algorithm 380: In-situ transposition of a rectangular
matrix, Comm. ACM IS, 5 (1970), 324-326
\bibitem{b15} ESKO G. CATE and DAVID W. TWIGG ALGORITHM 513: Analysis of In-Situ Transposition (1977)
\bibitem{b16} most of the crypto functions are written in assembly usin SIMD instructions in the
linux kernel "linux/arch/x86/crypto" for exanple transpose\_4x4 in "serpent-avx2-asm\_64.S", "sm4-aesni-avx-asm\_64.S",
"cast6-avx-x86\_64-asm\_64.S" and others
\bibitem{b17} https://github.com/torvalds/linux/commit/d34a460092d857f1616e39eed7eac6f40cea2225
\bibitem{b18} W. O. Alltop, "A Computer Algorithm for Transposing Nonsquare Matrices," in IEEE Transactions on Computers, vol. C-24, no. 10, pp. 1038-1040, Oct. 1975, doi: 10.1109/T-C.1975.224124.
\bibitem{b19} Jinwoo Suh and V. K. Prasanna, "An efficient algorithm for out-of-core matrix transposition," in IEEE Transactions on Computers, vol. 51, no. 4, pp. 420-438, April 2002, doi: 10.1109/12.995452.
\bibitem{b20} Kandemir, M. \& Choudhary, A. \& Ramanujam, J. \& Arunachalam, Meenakshi. (2000). A unified framework for optimizing locality, parallelism, and communication in out-of-core computations. Parallel and Distributed Systems, IEEE Transactions on. 11. 648 - 668. 10.1109/71.877759. 
\bibitem{b21} I-Jui Sung, Juan G´omez-Luna, Josè Marìa Gonzàlez-Linares
Nicolàs Gui, Wen-Mei W. Hwu. In-Place Transposition of Rectangular Matrices on Accelerators
\bibitem{b22} Post: An Efficient Matrix Transpose in CUDA C/C++. Feb 18, 2013 By Mark Harris, https://developer.nvidia.com/blog/efficient-matrix-transpose-cuda-cc/
\bibitem{b23} Hanounik, Bedros. (2000). Diagonal Registers: Novel Vector Register File Design for High Performance and Multimedia Computing. 
\bibitem{b24} https://github.com/San7o/valFuzz
\bibitem{b25} Rotenberg, A. (1960). "A New Pseudo-Random Number Generator". Journal of the ACM. 7 (1): 75–77
\bibitem{b26} Numerical Recipes ranqd1, Chapter 7.1, An Even Quicker Generator, Eq. 7.1.6
parameters from Knuth and H. W. Lewis
\bibitem{b27} https://github.com/San7o/tenno-tl
\end{thebibliography}
\vspace{12pt}
